---
title: 'EDS241: Assignment 2 Template'
author: "Charlie Curtin"
date: "format(Sys.time())"
output:
  html_document:
    toc: false
    df_print: paged
  pdf_document:
    toc: false
    number_sections: true
header-includes: \setlength{\parindent}{1em} \usepackage{float} \renewcommand{\thesubsection}{Question
  (\alph{subsection})}
--- 

**Reminders:** Make sure to read through the setup in markdown. Remember to fully report/interpret your results and estimates (in writing) + present them in tables/plots.
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
#Clean Environment
rm(list=ls())

# Setup your coding process in a way that works for you. Ideally use projects to organize your scripts and outputs. You all probably know more about this than us! For this project, I would create a project with all your data and scripts. I often store data on servers rather than my computer which is why I use the code you see below.

# I set an extension to retrieve data from a particular place (Google Drive/servers etc) and projects to organize my scripts and outputs on my computer/github.

# here I am setting a path to where I stored the data for this assignment
# data_wd <- "/Users/elliottfinn/Library/CloudStorage/GoogleDrive-elliottfinn@ucsb.edu/Shared drives/EDS241/Assignments/Assignment 2" 

# Example of how I use this Data Working Directory:
# data <- read_csv(paste0(data_wd,"/FILE_NAME.csv")) This helps me download/manage my data from different places.

# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)

# load packages
packages=c(
# Necessary for Assignment 2
  "Match","plm", "tidyverse", "MatchIt", "RItools", "Hmisc", "lmtest", "estimatr",
# You decide what works for you, these are the packages I use to display results ect, they may not be the ones you use.

"gridExtra", "stargazer", "kableExtra",
"purrr", "knitr", "broom",
   
  # Some Potentially useful packages from earlier examples
           "stargazer", "here", "tidyr", "dplyr","stringr", "janitor", 
           "cowplot", "ggplot2", "tinytex", "datasets", "tibble") # Used for Mock assignment

for (i in packages) {
  if (require(i,character.only=TRUE)==FALSE) {
    install.packages(i,repos='http://cran.us.r-project.org')
  }
  else {
    require(i,character.only=TRUE)
  }
}

# Disable scientific notation if you want
options(scipen=999)

```

# Part 1 Treatment Ignorability Assumption and Applying Matching Estimators (19 points):
The goal is to estimate the causal effect of maternal smoking during pregnancy on infant birth weight using the treatment ignorability assumptions. The data are taken from the National Natality Detail Files, and the extract “SMOKING_EDS241.csv”' is a random sample of all births in Pennsylvania during 1989-1991. Each observation is a mother-infant pair. The key variables are:

**The outcome and treatment variables are:**

\indent birthwgt=birth weight of infant in grams

\indent tobacco=indicator for maternal smoking

**The control variables are:**

\indent mage (mother's age), meduc (mother's education), mblack (=1 if mother identifies as Black), alcohol (=1 if consumed alcohol during pregnancy), first (=1 if first child), diabete (=1 if mother diabetic), anemia (=1 if mother anemic)

```{r , include = TRUE, message = FALSE, warning = FALSE}
# Load data for Part 1
smoking <- read_csv(here("data", "SMOKING_EDS241.csv"))
```

## Mean Differences, Assumptions, and Covariates _(3 pts)_
a) What is the mean difference in birth weight of infants with smoking and non-smoking mothers [1 pts]?  Under what assumption does this correspond to the average treatment effect of maternal smoking during pregnancy on infant birth weight [0.5 pts]? Calculate and create a table demonstrating the differences in the mean proportions/values of covariates observed in smokers and non-smokers (remember to report whether differences are statistically significant) and discuss whether this provides empirical evidence for or against this assumption. Remember that this is observational data. What other quantitative empirical evidence or test could help you assess the former assumption? [1.5 pts: 0.5 pts table, 1 pts discussion]

```{r , include=TRUE}
## Calculate mean difference. Remember to calculate a measure of statistical significance

## For binary variables you should use the proportions test
#prop.test()
# select binary variables
binary_vars <- smoking %>% 
  select(anemia, diabete, alcohol, mblack, first, tobacco)

binary_vars_names <- smoking %>% 
  select(anemia, diabete, alcohol, mblack, first) %>% 
  names()

# create a dataframe to store results
prop_test_results <- data.frame()

# create a function that runs a proportions test
for (var in binary_vars_names) {
  smoking_m <- binary_vars %>% 
    filter(tobacco == 1) %>% 
    pull(!!sym(var))
  nonsmoking_m <- binary_vars %>% 
    filter(tobacco == 0) %>% 
    pull(!!sym(var))
  prop_test_result <- prop.test(x = c(sum(smoking_m),
                                      sum(nonsmoking_m)),
                                n = c(length(smoking_m),
                                      length(nonsmoking_m)),
                                correct = FALSE)
  prop_test_result_tidy <- broom::tidy(prop_test_result)
  prop_test_result_tidy$Variable <- var
  prop_test_results <- rbind(prop_test_results, prop_test_result_tidy)
}


# for continuous variables
t_test_results <- data.frame()

cont_vars <- smoking %>% 
  select(mage, meduc, birthwgt, tobacco)

cont_vars_names <- smoking %>% 
  select(mage, meduc, birthwgt) %>% 
  names()

for (var in cont_vars_names) {
  formula <- as.formula(paste(var, "~ tobacco"))
  t_test_result <- t.test(formula, data = cont_vars)
  t_test_result_tidy <- broom::tidy(t_test_result) 
  t_test_result_tidy$Variable <- var
  t_test_results <- rbind(t_test_results, t_test_result_tidy)
}

# combine results into a table
combined_results <- bind_rows(
  prop_test_results %>% select(Variable, estimate1, estimate2, p.value), 
  t_test_results %>% select(Variable, estimate1, estimate2, p.value)
)

# format table
combined_results_table <- kable(combined_results, format = "html",
                                col.names = c("Variable", "Proportion or Mean Control",
                                              "Proportion or Mean Treated", "P-Value"),
                                caption = "Treated and Untreated Proportion and T-Test results summary") %>% 
  kable_styling(font_size = 7, latex_options = "hold_position")

## Covariate Calculations and Tables (feel free to use code from Assignment 1 key)
combined_results_table

# mean difference of birth weights
mean_diff <- (combined_results[8, 3] - combined_results[8, 2])
```

- the mean difference in birth weight between smoking and nonsmoking mothers is `r{mean_diff}`. This corresponds to the ATE under the treatment ignorability assumption, which means that means.  There are statistically significant differences between in the control variables between our treated and untreated groups. This provides empirical evidence against treatment ignorability since


## ATE and Covariate Balance _(3 pts)_
b) Assume that maternal smoking is randomly assigned conditional on the observable covariates listed above. Estimate the effect of maternal smoking on birth weight using an OLS regression with NO linear controls for the covariates [0.5 pts]. Perform the same estimate including the control variables [0.5 pts]. Next, compute indices of covariate imbalance between the treated and non-treated regarding these covariates (see example file from class). Present your results in a table [1 pts]. What do you find and what does it say regarding whether the assumption you mentioned responding to a) is fulfilled? [1 pts]

```{r, include = TRUE}
# ATE Regression univariate
model1 <- lm(birthwgt ~ tobacco, data = smoking)

# ATE with covariates
model2 <- lm(birthwgt ~ tobacco + anemia + diabete + alcohol + mblack + first + mage + meduc, data = smoking)

# calculate standard errors
se_models <- starprep(model1, model2, stat = c("std.error"), se_type = "HC2", alpha = .05)

# Present Regression Results
stargazer(model1, model2, se = se_models,
          type = "text", ci = FALSE, no.space = TRUE,
          header = FALSE, omit = c("Constant"), omit.stat = c("adj.rsq", "ser", "f"),
          covariate.labels = c("tobacco", "anemia", "diabete", "alcohol", "mblack", "first", "mage", "meduc"),
          dep.var.labels = c("Birthweight (g)"), dep.var.caption = c(""),
          title = "Average Treatment Effect of Smoking on Birthweight", table.placement = "H",
          notes = "Robust standard errors in parentheses", notes.align = "l")

# Covariate balance
cov_balance <- xBalance(tobacco ~ anemia + diabete + alcohol + mblack + first + mage + meduc, 
                        data = smoking, report = c("std.diffs", "chisquare.test", "p.values"))

# Balance Table 
# format table
tidy.xbal(cov_balance) %>% 
  select(-NA.info) %>% 
  kbl(caption = "covariate balance") %>% 
  kable_minimal()
```

- regression results- when you add all the covariates, the size of our estimate on tobacco changes. This means that there are other factors influencing 

- balance table- The standardized difference in the balance table. For p-values that are significant, we've rejected the null hypothesis that there is no difference between the proportions or means of the covariates in our treatment and control groups. It shows that all the covariates are imbalanced between our treatment and control groups besides diabetes. 
- The value .1 is the threshold of balance. Beneath it, they're balanced. The null hypothesis is that there is no difference, so a high p-value means that we 


## Propensity Score Estimation _(3 pts)_
c) Next, estimate propensity scores (i.e. probability of being treated) for the sample, using the provided covariates. Create a regression table reporting the results of the regression and discuss what the covariate coefficients indicate and interpret one coefficient [1.5 pts]. Create histograms of the propensity scores comparing the distributions of propensity scores for smokers ('treated') and non-smokers ('control'), discuss the overlap and what it means [1.5 pts].

```{r , include=TRUE}
## Propensity Scores
# using glm
propensity_lm <- glm(tobacco ~ anemia + diabete + alcohol + mblack + first + mage + meduc, 
                  data = smoking, family = binomial())

summary(propensity_lm)

# predict propensity scores
smoking$pscores <- predict(propensity_lm, type = "response")


## PS Histogram Unmatched 
# histogram of propensity scores before matching
histbackback(split(smoking$pscores,	smoking$tobacco),	main = 
  "Propensity	score	before matching",	xlab = c("control",	"treatment"))

# format regression table
stargazer(propensity_lm, type = "text")
```

- regression results- The covariate coefficients show the increase in log-odds of the mother being a smoker (selecting into treatment), holding all other variables constant. For anemia, if the mother has anemia, their log-odds of being a smoker increase by .334. 

-histograms- The overlap in the histograms between control and treatment show that we do have matches at every bin of propensity scores, so we can try matching. 



## Matching Balance _(3 pts)_
(d) Next, match treated/control mothers using your estimated propensity scores and nearest neighbor matching. Compare the balancing of pretreatment characteristics (covariates) between treated and non-treated units in the original dataset (from c) with the matched dataset (think about comparing histograms/regressions) [2 pts]. Make sure to report and discuss the balance statistics [1 pts].

```{r, include = TRUE}
## Nearest-neighbor Matching
m.nn	<- matchit(tobacco ~ anemia + diabete + alcohol + mblack + first + mage + meduc,
                data = smoking,	method = "nearest",	ratio = 1)
summary(m.nn)

match.data = match.data(m.nn)

## Covariate Imbalance post matching: 
cov_balance_ps <- xBalance(tobacco ~ anemia + diabete + alcohol + mblack + first + mage + meduc, 
                           data = match.data, report = c("std.diffs", "chisquare.test", "p.values"))

tidy.xbal(cov_balance_ps) %>% 
  select(-NA.info) %>% 
  kbl(caption = "covariate balance") %>% 
  kable_minimal()

## Histogram of PS after matching
histbackback(split(match.data$distance, match.data$tobacco),	main = 
  "Propensity	score	after matching",	xlab = c("control",	"treatment"))
```
- histograms- Our histogram after matching shows more balance between control and treatment groups. We matched observations in the treatment and control groups based on their propensity scores. 

-balanced statistics- now we have matched on all the covariates except for alcohol and meduc. That means using nearest neighbors, it's difficult to find a good counterfactual for mothers in the treatment group who smoke. Same for meduc. Our high p-value indicates that our other covariates are balanced because the std difference is under the .1 threshold, and they are statistically insignificant. 


## ATT with Nearest Neighbor _(3 pts)_
(e) Estimate the ATT using the matched dataset. Report and interpret your result (Note: no standard error or significance test is required here)

```{r , include=TRUE}
## Nearest Neighbor 
## ATT
sumdiff_data <- match.data %>%
  group_by(subclass) %>%
  mutate(diff = birthwgt[tobacco == 1] - birthwgt[tobacco == 0])

# find ATT using the matched dataset
sumdiff <- sum(sumdiff_data$diff) / 2
ATT_m_nn = 1 / sum(smoking$tobacco) * sumdiff
ATT_m_nn

# ATT is also the mean of the difference
mean(sumdiff_data$diff)
```

- The average treatment effect on the treated means that for mothers who smoke during pregnancy, their babies are, on average, born 222.94 grams lighter than mothers who don't smoke during pregnancy. 


## ATE with WLS Matching _(3 pts)_
f) Last, use the original dataset and perform the weighted least squares estimation of the ATE using the propensity scores (including controls). Report and interpret your results, here include both size and precision of estimate in reporting and interpretation.

```{r , include=TRUE}
## Weighted least Squares (WLS) estimator Preparation
ps <- smoking$pscores # define birth weights
Y <- smoking$birthwgt # define our outcome variable
D <- smoking$tobacco # define our treatment variable

# calculation of the weights based on propensity scores
smoking$wgt = (D / ps + (1 - D) / (1 - ps))

## Weighted least Squares (WLS) Estimates
wls_controls	<-lm(birthwgt	~ tobacco	+ anemia + diabete + alcohol + mblack + first + mage + meduc,
                  data = smoking, weights = wgt)

summary(wls_controls)

## Present Results
# format table

```

- size = coefficient
- precision = standard error, confidence interval, significance level
- Using least weighted squares regression, we estimate that the difference in mean birth weight between smoking and nonsmoking mothers is -224.85 grams. With WLS, we're using the entire dataset. With nearest neighbor matching, we're dropping lots of data.


## Differences in Estimates _(1 pts)_ 
g) Explain why it was to be expected given your analysis above that there is a difference between your estimates in e) and f)? 

There's only a slight difference between the estimates of the ATT (using nearest neighbor matching) and ATE (using weighted least squares regression). 


\newpage

# Part 2 Panel model and fixed effects (6 points)
\indent We will use the  progresa data from last time as well as a new dataset. In the original dataset, treatment households had been receiving the transfer for a year. Now, you get an additional dataset with information on the same households from before the program was implemented, establishing a baseline study (from 1997), and the same data we worked with last time (from 1999).
\indent *Note: You will need to install the packages plm and dplyr (included in template preamble). Again, you can find a description of the variables at the bottom of PDF and HERE.

## Estimating Effect with First Difference _(3 pts: 1.5 pts estimate, 1.5 pts interpretation)_
Setup: Load the new baseline data (progresa_pre_1997.csv) and the follow-up data (progresa_post_1999.csv) into R. Note that we created a time denoting variable (with the same name, 'year') in BOTH datasets. Then, create a panel dataset by appending the data (i.e. binding the dataset row-wise together creating a single dataset). We want to examine the same outcome variable as before, value of animal holdings (vani).

```{r , include=TRUE}
rm(list=ls()) # clean environment

## Load the datasets
progresa_post_1999 <- read_csv(here("data", "progresa_post_1999.csv"))
progresa_pre_1997 <- read_csv(here("data", "progresa_pre_1997.csv"))

## Append post to pre dataset 
progresa <- rbind(progresa_pre_1997, progresa_post_1999)
```
a) Estimate a first-difference (FD) regression manually, interpret the results briefly (size of coefficient and precision!)
\indent *Note: Calculate the difference between pre- and post- program outcomes for each family. To do that, follow these steps and the code given in the R-template:

```{r, include=TRUE}
### Code included to help get you started
## i. Sort the panel data in the order in which you want to take differences, i.e. by household and time.

## Create first differences of variables
 progresa <- progresa %>% 
   arrange(hhid, year) %>% 
   group_by(hhid) %>% 

## ii. Calculate the first difference using the lag function from the dplyr package.
   mutate(vani_fd = vani - dplyr::lag(vani))

## iii. Estimate manual first-difference regression (Estimate the regression using the newly created variables.)
fd_manual <- lm(vani_fd ~ treatment, data = progresa)

summary(fd_manual)

```
- ATE of the cash flow on value of animal holdings is 287.90 using a first difference estimator

## Fixed Effects Estimates _(2 pts: 1 pts estimate, 1.5 interpretation)_
b) Now also run a fixed effects (FE or ‘within’) regression and compare the results. Interpret the estimated treatment effects briefly (size of coefficient and precision!)

```{r, include=TRUE}
## Fixed Effects Regression
within1 <- plm(vani ~ treatment, index = c("state", "year"), model = "within", effect = "twoways", data = progresa)

## Present Regression Results
summary(within1)

# replace NA values with 0s
```

- fixed effects control for differences between groups. In our example, we're controlling for differences between states. We're worried that there might be differences between states. In fixed effects you compare the effects of the group to itself over time, 


## First Difference and Fixed Effects and Omitted Variable Problems _(1 pts)_
c) Explain briefly how the FD and FE estimator solves a specific omitted variable problem? Look at the example on beer tax and traffic fatalities from class to start thinking about ommitted variables. Give an example of a potential omitted variable for the example we are working with here that might confound our results? For that omitted variable, is a FE or FD estimator better? One example is enough.

